---
title: "WKSANDEEL - read in Norwegian sample data"
author: "Kirsten Birch Håkansson, DTU Aqua"
date: \today
output: pdf_document
---

```{r setup, include=FALSE}
library(dplyr)
library(lubridate)
library(haven)
library(ggplot2)
library(RColorBrewer)
library(knitr)
library(tidyverse)

options(scipen = 999)
knitr::opts_chunk$set(fig.width = 8, fig.height = 8, dpi = 900, dev = 'jpeg', results = 'asis', echo = F, warning = F)

path_old <- "../boot/data/old_nor_data/"
path_new <- "../boot/data/"
path_out <- "../data/"

```


# Purpose

Errors was discovered in the Norwegian data submitted to HAWG the last couple of year. Often there were more length classes in the ALK than the length distribution.

The same error seems to be present in all years in the old time series

Norway has submitted data for the whole time period. Newly extracted data back to 2010 and a old extract of data from to 2009.

The two timeries are in different formats

scm: this is cm in norwegian_alk.sas7bdat & norwegian_length_freq.sas7bdat

*OBS - all plots states scm, but it is cm*

stat_unique: stat, ICESsq, month, day - added unique stat number. stat already in the dataset is not unique. 

PP_ar_tx: not added - it is added later in the SAS scripts and it don't really fits in the RDB format.

month: just normal month - it looks like half month and so on are added in the SAS scripts and it don't really fits in the RDB format.

RDB format:

This is not a full set of CS files, since submitted info are scarce - an HH, SL, HL and CA file is outputtet with a lot of NA's

**TODO**

Check for duplicated ld's (not in 2018). When runnning a distinct on HL, then around 1500 rows are deleted.

# Lengths
## Read in data in old and new format, and combine


```{r}

# Data from 2024 - assessment 2025
ld_new_02 <- read.csv(paste0(path_new, "Annex_1_HAWG_sandeel_exchange_format_Norway_Table2.csv"))

ld_new_02$description <- "LD - Data extracted in 2025"

names(ld_new_02)

length(unique(ld_new_02$SampleId))
ld_new_02$SampleId <- paste(ld_new_02$year, ld_new_02$SampleId, sep = "-")

length(unique(ld_new_02$SampleId))
length(unique(paste(ld_new_02$SampleId)))
length(unique(paste(ld_new_02$year, ld_new_02$SampleId, ld_new_02$samplingDate)))

ld_mw_02 <- subset(ld_new_02, !(is.na(weightAtLengthClass)))

unique(ld_new_02$year)

# Data from 2023 - assessment 2024
ld_new_01 <- read.csv(paste0(path_old, "Annex_1_HAWG_sandeel_exchange_format_Norway_Table2_2023.csv"))

ld_new_01$description <- "LD - Data extracted in 2024"

names(ld_new_01)

length(unique(ld_new_01$SampleId))
ld_new_01$SampleId <- paste(ld_new_01$year, ld_new_01$SampleId, sep = "-")

length(unique(ld_new_01$SampleId))
length(unique(paste(ld_new_01$SampleId)))
length(unique(paste(ld_new_01$year, ld_new_01$SampleId, ld_new_01$samplingDate)))

ld_mw_01 <- subset(ld_new_01, !(is.na(weightAtLengthClass)))

unique(ld_new_01$year)

# Data from 2022 - assessment 2023
ld_new_0 <- read.csv(paste0(path_old, "Annex_1_HAWG_sandeel_exchange_format_Norway_Table2_2022.csv"))

ld_new_0$description <- "LD - Data extracted in 2023"

names(ld_new_0)

length(unique(ld_new_0$SampleId))
ld_new_0$SampleId <- paste(ld_new_0$year, ld_new_0$SampleId, sep = "-")

length(unique(ld_new_0$SampleId))
length(unique(paste(ld_new_0$SampleId)))
length(unique(paste(ld_new_0$year, ld_new_0$SampleId, ld_new_0$samplingDate)))

ld_mw_0 <- subset(ld_new_0, !(is.na(weightAtLengthClass)))

unique(ld_new_0$year)


# Data from 2007-2021 benchmark data
ld_new <- read.csv(paste0(path_old, "Annex_1_HAWG_sandeel_exchange_format_Norway_Table2_2007_2021.csv"))

ld_new$description <- "LD - Data extracted in 2022"

names(ld_new)

length(unique(ld_new$SampleId))
ld_new$SampleId <- paste(ld_new$year, ld_new$SampleId, sep = "-")

length(unique(ld_new$SampleId))
length(unique(paste(ld_new$SampleId)))
length(unique(paste(ld_new$year, ld_new$SampleId, ld_new$samplingDate)))

ld_mw <- subset(ld_new, !(is.na(weightAtLengthClass)))

unique(ld_new$year)

# combine new

ld_mw <- rbind(ld_mw, ld_mw_0, ld_mw_01, ld_mw_02)
ld_new <- rbind(ld_new, ld_new_0, ld_new_01, ld_new_02)

# Old
ld_old_ld <-
  read.delim(paste(path_old, "Norwegian_sandeel_length_freq_79_09.txt", sep = ""))

ld_old_alk <-
  read.delim(paste(path_old, "Norwegian_ALK_data_97_09.txt", sep = ""))

names(ld_old_ld)
names(ld_old_alk)

ld_old <- bind_rows(ld_old_ld, ld_old_alk)

ld_old$date <-
  as.Date(str_pad(as.character(ld_old$date), 6, "left", 0), "%y%m%d")

unique(ld_old$date)
str(ld_old)

# Transform into new format

ld_old$vesselFlagCountry <- "NO"
ld_old$year <- year(ld_old$date)

plot(ld_old$year, ld_old$date)

ld_old$SampleId <-
  paste(ld_old$year,
        ld_old$Station,
        ld_old$trip,
        ld_old$Rectangle,
        ld_old$date,
        sep = "-")

# Test uniqueness of sampledId
length(unique(paste0(ld_old$Station, ld_old$trip)))
length(unique(ld_old$SampleId))
unique(ld_old$Station)
unique(ld_old$trip)

ld_old$samplingDate <- ld_old$date
ld_old$area[ld_old$Area == "N"] <- "27.4"
ld_old$area[ld_old$Area == "S"] <- "27.3.a.20"


unique(ld_old$area)
distinct(ld_old, area, Area)

ld_old$statisticalRectangle <- ld_old$Rectangle

unique(ld_old$Rectangle)

ld_old$exclusiveEconomicZone <- NA
ld_old$lengthClass <- (ld_old$length_SCM / 2) * 10

hist(ld_old$length_SCM, )  # original length
hist(ld_old$lengthClass) # length in mm

ld_old$numberAtLengthClass <- ld_old$Total_number_length_class
ld_old$weightAtLengthClass <- ld_old$Total_weight_length_class

ld_old$description <- "LD - Data extracted in the past"

# Select only needed fields

unique(ld_old$year)
unique(ld_new$year)

ld_old_1 <- select(ld_old, one_of(names(ld_new)))
ld_old_2 <- subset(ld_old_1, year < min(ld_new$year))

ld <- rbind(ld_old_2, ld_new)
distinct(ld, year, description)

ld_sum <- summarise(group_by(ld, year), no_fish = sum(numberAtLengthClass, na.rm = T),
                                                      no_samples = length(unique(SampleId)))

kable(ld_sum)
```


# ALK
## Read in data in old and new format, and combine

```{r}

# Data from 2024 - assessment 2025

alk_new_02 <- read.csv(paste0(path_new, "Annex_1_HAWG_sandeel_exchange_format_Norway_Table3.csv"))

alk_new_02$description <- "ALK - Data extracted in 2025"

alk_new_02$plusGrp <- NA

alk_new_02$SampleId <- paste(alk_new_02$year, alk_new_02$SampleId, sep = "-")

length(unique(alk_new_02$SampleId))
length(unique(paste(alk_new_02$year, alk_new_02$SampleId)))
length(unique(paste(alk_new_02$year, alk_new_02$SampleId, alk_new_02$samplingDate)))

# Data from 2023 - assessment 2024

alk_new_01 <- read.csv(paste0(path_old, "Annex_1_HAWG_sandeel_exchange_format_Norway_Table3_2023.csv"))

alk_new_01$description <- "ALK - Data extracted in 2024"

alk_new_01$plusGrp <- NA

alk_new_01$SampleId <- paste(alk_new_01$year, alk_new_01$SampleId, sep = "-")

length(unique(alk_new_01$SampleId))
length(unique(paste(alk_new_01$year, alk_new_01$SampleId)))
length(unique(paste(alk_new_01$year, alk_new_01$SampleId, alk_new_01$samplingDate)))

# Data from 2022 - assessment 2023

alk_new_0 <- read.csv(paste0(path_old, "Annex_1_HAWG_sandeel_exchange_format_Norway_Table3_2022.csv"))

alk_new_0$description <- "ALK - Data extracted in 2023"

alk_new_0$plusGrp <- NA

alk_new_0$SampleId <- paste(alk_new_0$year, alk_new_0$SampleId, sep = "-")

length(unique(alk_new_0$SampleId))
length(unique(paste(alk_new_0$year, alk_new_0$SampleId)))
length(unique(paste(alk_new_0$year, alk_new_0$SampleId, alk_new_0$samplingDate)))

# Data from 2007-2021 benchmark data

alk_new <- read.csv(paste0(path_old, "Annex_1_HAWG_sandeel_exchange_format_Norway_Table3_2007_2021.csv"))

alk_new$description <- "ALK - Data extracted in 2022"

alk_new$plusGrp <- NA

alk_new$SampleId <- paste(alk_new$year, alk_new$SampleId, sep = "-")

length(unique(alk_new$SampleId))
length(unique(paste(alk_new$year, alk_new$SampleId)))
length(unique(paste(alk_new$year, alk_new$SampleId, alk_new$samplingDate)))

# combine new

alk_new <- rbind(alk_new, alk_new_0, alk_new_01, alk_new_02)

# Old

alk_old <-
  read.delim(paste(path_old, "Norwegian_ALK_data_97_09.txt", sep = ""))

alk_old$date <-
  as.Date(str_pad(as.character(alk_old$date), 6, "left", 0), "%y%m%d")

names(alk_old)

# Transform into new format

alk_old$vesselFlagCountry <- "NO"
alk_old$year <- year(alk_old$date)

plot(alk_old$year, alk_old$date)

alk_old$SampleId <-
  paste(alk_old$year,
        alk_old$Station,
        alk_old$trip,
        alk_old$Rectangle,
        alk_old$date,
        sep = "-")

# Test uniqueness of sampledId
length(unique(paste0(alk_old$Station, alk_old$trip)))
length(unique(alk_old$SampleId))
unique(alk_old$Station)
unique(alk_old$trip)

alk_old$samplingDate <- alk_old$date
alk_old$area[alk_old$Area == "N"] <- "27.4"
alk_old$area[alk_old$Area == "S"] <- "27.3.a.20"

unique(alk_old$area)
distinct(alk_old, area, Area)

alk_old$statisticalRectangle <- alk_old$Rectangle

unique(alk_old$Rectangle)

alk_old$exclusiveEconomicZone <- NA
alk_old$fishID <- NA
alk_old$lengthClass <- (alk_old$length_SCM / 2) * 10

hist(alk_old$length_SCM)  # original length
hist(alk_old$lengthClass) # length in mm

alk_old$weight <- alk_old$Mean_weight

plot(alk_old$lengthClass, alk_old$weight)
plot(alk_new$lengthClass, alk_new$weight)

alk_old$a0 <- alk_old$Age0
alk_old$a1 <- alk_old$Age1
alk_old$a2 <- alk_old$Age2
alk_old$a3 <- alk_old$Age3
alk_old$a4 <- alk_old$Age4.
alk_old$plusGrp <- 4

alk_old$description <- "ALK - Data extracted in the past"

alk_old_t <-
  filter(mutate(
    gather(alk_old, key = age, value = antal, a0, a1, a2, a3, a4),
    age = substr(age, 2, 2)
  ), antal != 0)

# Test transpose

sum(alk_old_t$antal, na.rm = T)
sum(alk_old$a0 + alk_old$a1 + alk_old$a2 + alk_old$a3 + alk_old$a4)

# Select only needed fields

alk_old_t_1 <- select(alk_old_t, one_of(names(alk_new)), antal)

# Add a line per fish
alk_old_t_2 <-
  alk_old_t_1[rep(row.names(alk_old_t_1), alk_old_t_1$antal),]

nrow(alk_old_t_2)

alk_old_done <- select(alk_old_t_2, -antal)

# Add weights from LD

ld_mw$weight <- ld_mw$weightAtLengthClass/ld_mw$numberAtLengthClass
ld_mw$fishID <- NA
ld_mw$age <- NA
ld_mw$plusGrp <- NA

ld_mw_1 <-
  ld_mw[rep(row.names(ld_mw), ld_mw$numberAtLengthClass),]

sum(ld_mw$numberAtLengthClass)
nrow(ld_mw_1)

ld_mw_2 <- select(ld_mw_1, one_of(names(alk_new)))

# Combine
unique(alk_old_done$year)
unique(alk_new$year)
unique(ld_mw_2$year)

alk_old_done_1 <- subset(alk_old_done, year < min(alk_new$year))

alk <- rbind(alk_old_done_1, alk_new, ld_mw_2)

names(alk)

distinct(alk, year, description)

alk_sum <- summarise(group_by(alk, year), no_fish = length(fishID),
                                                      no_samples = length(unique(SampleId)))

kable(alk_sum)
```


```{r}
ggplot(alk, aes(x = year, fill = description)) +
  geom_histogram(binwidth = 1)

```


```{r}
ggplot(ld, aes(x = year, fill = description)) +
  geom_histogram(binwidth = 1)

```


```{r}
ggplot(alk, aes(x = lengthClass, y = weight, col = description)) +
  geom_point()
```

```{r}
ggplot(alk, aes(x = lengthClass, col = description)) +
  geom_histogram()

```


```{r}
ggplot(alk, aes(x = month(samplingDate), col = description)) +
  geom_histogram()

```

# Output data in RDB format

```{r}

# HH

hh_names <- select(read.csv(paste("Q:/dfad/users/kibi/data/RDB/RDB_in_csv/rHH_2019.csv", sep = "")), -domain_dvh90119, -gear, -domain_maske)
hh_names$trpCode <- as.character(hh_names$trpCode)
hh_names$staNum <- as.character(hh_names$staNum)
hh_names$date <- as.Date(hh_names$date)

length(unique(alk$SampleId))

alk_hh <- select(alk, year, samplingDate, SampleId, area,
    statisticalRectangle)

ld_hh <- select(ld, year, samplingDate, SampleId, area,
    statisticalRectangle)

dat_hh <- bind_rows(alk_hh, ld_hh)

hh_nor <-
  distinct(
    dat_hh,
    recType = "HH",
    sampType = NA,
    landCtry = NA,
    vslFlgCtry = "NOR",
    year = year,
    quarter = quarter(samplingDate),
    month = month(samplingDate),
    proj = "NOR-sandeel",
    trpCode = SampleId,
    staNum = SampleId,
    foVal = "V",
    aggLev = NA,
    catReg = NA,
    sppReg = NA,
    date = samplingDate,
    time = NA,
    foDur = NA,
    latIni = NA,
    lonIni = NA,
    latFin = NA,
    lonFin = NA,
    area = area,
    rect = statisticalRectangle,
    subRect = NA,
    foDep = NA,
    waterDep = NA,
    foCatNat = NA,
    foCatEu5 = NA,
    foCatEu6 = NA,
    meshSize = NA,
    selDev = NA,
    meshSizeSelDev = NA
  )

hh_nor_1 <- select(hh_nor, one_of(names(hh_names)))
hh_nor_2 <- filter(bind_rows(hh_names, distinct(hh_nor_1)), is.na(sampType))

saveRDS(hh_nor_2, paste(path_out, "rdb_hh_nor.rds", sep = ""))
write.csv(hh_nor_2, paste(path_out, "rdb_hh_nor.csv", sep = ""))

# SL
sl_names <- read.csv(paste("Q:/dfad/users/kibi/data/RDB/RDB_in_csv/rSL_2019.csv", sep = ""))
sl_names$trpCode <- as.character(sl_names$trpCode)
sl_names$staNum <- as.character(sl_names$staNum)

sl_nor <-
  mutate(
    ld,
    recType = "SL",
    sampType = NA,
    landCtry = NA,
    vslFlgCtry = "NOR",
    year = year,
    quarter = quarter(samplingDate),
    month = month(samplingDate),
    proj = "NOR-sandeel",
    trpCode = SampleId,
    staNum = SampleId,
    sppName = "Ammodytes marinus",
    catchCat = NA,
    landCat = NA,
    commCatScl = NA,
    commCat = NA,
    subSampCat = NA,
    sex = NA,
    wt = NA,
    subSampWt = NA,
    lenCode = "scm",
    sampNo = NA,
    subSampNo = NA,
    wt_treatment = NA,
    subSampWt_treatment =NA
  )

sl_nor_1 <- select(sl_nor, one_of(names(sl_names)))
sl_nor_2 <- filter(bind_rows(sl_names, distinct(sl_nor_1)), is.na(sampType))

saveRDS(sl_nor_2, paste(path_out, "rdb_sl_nor.rds", sep = ""))
write.csv(sl_nor_2, paste(path_out, "rdb_sl_nor.csv", sep = ""))

#HL
hl_names <- read.csv(paste("Q:/dfad/users/kibi/data/RDB/RDB_in_csv/rHL_2019.csv", sep = ""))
hl_names$trpCode <- as.character(hl_names$trpCode)
hl_names$staNum <- as.character(hl_names$staNum)

hl_nor <-
  mutate(
    ld,
    recType = "HL",
    sampType = NA,
    landCtry = NA,
    vslFlgCtry = "NOR",
    year = year,
    quarter = quarter(samplingDate),
    month = month(samplingDate),
    proj = "NOR-sandeel",
    trpCode = SampleId,
    staNum = SampleId,
    sppName = "Ammodytes marinus",
    catchCat = NA,
    landCat = NA,
    commCatScl = NA,
    commCat = NA,
    subSampCat = NA,
    sex = NA,
    indSex = NA,
    lenCls = lengthClass,
    lenNum = numberAtLengthClass,
    sampLenNum = numberAtLengthClass
  )

hl_nor_1 <- select(hl_nor, one_of(names(hl_names)))
hl_nor_2 <- filter(bind_rows(hl_names, hl_nor_1), is.na(sampType))

saveRDS(hl_nor_2, paste(path_out, "rdb_hl_nor.rds", sep = ""))
write.csv(hl_nor_2, paste(path_out, "rdb_hl_nor.csv", sep = ""))


# CA

ca_names <-
  read.csv(paste("Q:/dfad/users/kibi/data/RDB/RDB_in_csv/rCA_2019.csv", sep = ""),
           nrow = 5000)
ca_names$trpCode <- as.character(ca_names$trpCode)
ca_names$staNum <- as.character(ca_names$staNum)

rownames(alk) = seq(length=nrow(alk))

ca_nor <-
  mutate(
    alk,
    recType = "CA",
    sampType = NA,
    landCtry = NA,
    vslFlgCtry = "NOR",
    year = year,
    quarter = quarter(samplingDate),
    month = month(samplingDate),
    proj = "NOR-sandeel",
    trpCode = SampleId,
    staNum = SampleId,
    area = area,
    rect = statisticalRectangle,
    subRect = NA,
    stock = NA,
    sppName = "Ammodytes marinus",
    catchCat = NA,
    landCat = NA,
    commCatScl = NA,
    commCat = NA,
    sex = NA,
    lenCls = lengthClass,
    age = as.numeric(age),
    fishId = as.numeric(row.names(alk)),
    lenCode = "scm",
    ageMeth = NA,
    plusGrp = as.character(plusGrp),
    otoWt = NA,
    otoSide = NA,
    indWt = weight,
    matMeth = NA,
    matScale = NA,
    matStage = NA,
    otolithReadingRemark = NA
  )

ca_nor_1 <- select(ca_nor, one_of(names(ca_names)))
ca_nor_2 <- filter(bind_rows(ca_names, ca_nor_1), is.na(sampType))

saveRDS(ca_nor_2, paste(path_out, "rdb_ca_nor.rds", sep = ""))
write.csv(ca_nor_2, paste(path_out, "rdb_ca_nor.csv", sep = ""))

# Testing join

hhhl <- left_join(hl_nor_2, select(hh_nor_2, -recType, -sampType))

hhhl_1 <- inner_join(hl_nor_2, select(hh_nor_2, -recType, -sampType))

hl_ca <- left_join(hl_nor_2, select(ca_nor_2, -recType, -sampType))
hl_ca_1 <- inner_join(hl_nor_2, select(ca_nor_2, -recType, -sampType))

```

```{r}


alk_2006 <- subset(alk, year == 2006)
length(unique(alk_2006$SampleId))
length(alk_2006$SampleId)


ld_2006 <- subset(ld, year == 2006)
length(unique(ld_2006$SampleId))
length(ld_2006$SampleId)
sum(ld_2006$numberAtLengthClass)


alk_2010_a <- subset(alk, year == 2010 & !(is.na(age)))
length(unique(alk_2010_a$SampleId))

ld_2010 <- subset(ld, year == 2010)
length(unique(ld_2010$SampleId))

alk_2010 <- subset(alk, year == 2010)
length(unique(alk_2010$SampleId))


```

